---
title: "Mixed models"
author: "Filippo Biscarini"
date: "12/31/2021"
output: 
  html_document:
    keep_md: false
  github_document:
    pandoc_args: ['--lua-filter', 'math-github.lua']
bibliography: ../bibliography/biostatistics.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("katex")
```

## Mixed models

Linear mixed models are a very powerful statistical tool that, simply put, allows to mix "fixed" (systematic) and "random" effects in one model equation:  

```{r, echo=FALSE}
# tex = "\\mathbf{y} = \\mathbf{Xb} + \\mathbf{Zu} + \\mathbf{e}"
# katex_html(
#   tex,
#   displayMode = TRUE,
#   include_css = TRUE,
#   preview = interactive()
# )
```

$$
\mathbf{y} = \mathbf{Xb} + \mathbf{Zu} + \mathbf{e}
$$

- **y**: vector of response (target) variable (observations) $(n,1)$
- **b**: vector of fixed effects $(f,1)$
- **u**: vector of random effects $(r,1)$
- **e**: vector of model residuals $(n,1)$
- **X**: design matrix that relates observations (responses) with fixed effects $(n,f)$
- **Z**: design matrix that relates observations (responses) with random effects $(n,r)$

The *letters* between brackets refer to the vector/matrix dimensions: *n* for the number of records (observations), *f* for the number of fixed effects, *r* for the number of random effects.

This is a **linear additive mixed model**: 

- *linear* because it is a linear combination of variables (a.k.a. "parameters"): **b**, **u** (and *e*)
- *additive* because variables are summed together
- *mixed* because (again) it mixes fixed and random effects

While **fixed effects** are just point estimates of underlying *true* values, **random effects** are associated with a distribution and, importantly, a **variance component**. 
The total variance of the observations (**y**'s) is partitioned into one component belonging to the random effect **u** and the residual variance:

```{r, echo=FALSE}
# tex = "Var(\\mathbf{y}) = \\mathbf{K} \\sigma_u^2 + \\mathbf{R} + \\sigma_e^2"
# katex_html(
#   tex,
#   displayMode = TRUE,
#   include_css = TRUE,
#   preview = interactive()
# )
```

$$
Var(\mathbf{y}) = \mathbf{K} \sigma_u^2 + \mathbf{R} \sigma_e^2
$$

where:

- **K**: *covariance* (relationship) *matrix* between the observations (for the specific random effect)
- **R** is the residual covariance matrix, usually simplified to $\mathbf{I}$ (identity matrix) under the assumption of independence between residuals

Estimates for the parameters in the mixed model above are obtained by solving the **mixed model equations** (MME, @henderson1950estimaton). MME are set up by expanding the equation $\left(\mathbf{X'+Z'}\right) \cdot \left(\mathbf{Xb}+ \mathbf{Zu} \right) = \left( \mathbf{X'+Z'} \right) \cdot \mathbf{y}$, and then adding to the random term **u** the corresponding covariance matrix:

$$
\left[
\begin{array}{cc}
\mathbf{X'X} & \mathbf{X'Z} \\
\mathbf{Z'X} & \mathbf{Z'Z} + \mathbf{K^{-1}\cdot \frac{\sigma_e^2}{\sigma_u^2}}
\end{array}
\right] \cdot
\left[ 
\begin{array}{c}
\mathbf{b}\\
\mathbf{u}
\end{array}
\right]
= \left[
\begin{array}{c}
\mathbf{X'y}\\
\mathbf{Z'y}
\end{array}
\right]
$$

**MMEs** are a very powerful statistical/mathematical tool, that can be extended in a very flexible way to a very large number of different (and increasingly) models (see for instance @mrode2014linear)

### Mixed models: an illustration in R

We take data on the yield of 41 potato lines (`DT_example` dataset) from different environments (combination of location and year).
There are *185* records.

```{r mixed_model, warning=FALSE, message=FALSE}
library("sommer")
library("tidyverse")
data(DT_example)
DT <- DT_example

head(DT)
```

We first look at the distribution of the quantitative response variable:

```{r}
hist(DT$Yield, xlab = "yield", main="response variable")
```
Then at the distribution of observations (yield) per environemnt:

```{r}
p <- ggplot(DT, aes(x = Name, y = Yield)) + geom_jitter(aes(color=Name)) + facet_wrap(~Env)
p <- p + theme(legend.position = "none", axis.text.x = element_blank())
p
```

### Fitting the mixed model

We now fit a mixed model of the form:

```{r, echo=FALSE}
# tex = "\\mathbf{yield} = \\mathbf{X} \\cdot \\mathbf{env} + \\mathbf{Z} \\cdot \\mathbf{variety} + \\mathbf{e}"
# katex_html(
#   tex,
#   displayMode = TRUE,
#   include_css = TRUE,
#   preview = interactive()
# )
```

$$
\mathbf{yield} = \mathbf{X} \cdot \mathbf{env} + \mathbf{Z} \cdot \mathbf{variety} + \mathbf{e}
$$

where we take the *environment* as a "fixed" systematic effect to be point-estimated, and *variety* as the random effect of the potato varieties (a proxy for a *genetic effect*, but with no specified kinship matrix).
This model assumes one variance component for the random effect *variety* (variable `Name` in the code), and one residual variance component (`rcov = ~ units`, in the code):

```{r}
fit <-mmer(fixed = Yield ~ Env, 
            random = ~ Name,
            rcov = ~ units,
            data = DT, 
            verbose = FALSE)
```

```{r}
summary(fit)
```

```{r}
ratio <- fit$sigmaVector[1]/fit$sigmaVector[2]
fit$sigmaVector
```

From the object returned by the `mmer` function, we can see that the variance components are:

- $\sigma_u^2$ = `r fit$sigmaVector[1]`, for the **variety** variance component
- $\sigma_e^2$ = `r fit$sigmaVector[2]`, for the **residual** variance component

This gives a ratio of `r round(ratio,3)` for the proportion of total phenotypic variance (*yield*) explained by the **variety** effect, which corresponds to **`r round(ratio,3)*100`%**.

## References

